---
title: "Subpopulation analysis of race and ethnicity in complex survey data"
author: "Nathan Alexander"
output: 
  html_document:
    self_contained: yes
editor_options: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Use outline and rework content for intro sections from [here](https://www.dropbox.com/scl/fi/fbiquhrn7o1jcwl0bl4o4/Mind-the-Gap-IUME-December-5-2013_final.pptx?rlkey=mm6wel3wyn79nyekdms23y4dq&dl=0)

## Abstract

Research on subpopulation data in complex survey designs is a common
practice in statistics. Despite a wealth of resources on the issues in
the measurement and standard error inaccuracies in subset population
data, few studies provide an understanding of measurement in the context
of group-based outcomes. Decades of research on eugenics and racialized
statistical methodologies has set a solid foundation to the issue of
measuring race in statistics. This critique includes important reminders
about the philosophy, theory, bias, and measurement issues of
\`\`race-based" analyses in statistics. The central role of statistics
and measurement in modern policy discourse around race-based group
outcomes, however, has not changed. As new frameworks around critical
inquiry and statistical methods continue to expand, this article
provides a review of the literature and measurement components of race
and ethnicity in complex survey data. We discuss implications in
relation to racialization and the construction of subpopulations,
subgroups, and domains in complex survey data. We use data from two U.S.
nationally-represntative complex surveys studies and provide
implications for policy and researcher practice.

#### Keywords

Subgroup, domain, variance, survey, complex design

# Overview

This document details some technical foundations and data related to the
analysis of a sub-population in complex survey data.

If $p$ is the proportion who are Black and $(1-p)$ is the proportion who
are not Black, and $$ \bar{X} = p\bar{X}_d + (1-p)\bar{X}_a $$
<https://stats.stackexchange.com/questions/30562/how-to-test-whether-subgroup-mean-differs-from-overall-group-that-includes-the>

### Race

The minimum categories established in the OMB 1997 Standard are

-   Race

    -- American Indian or Alaska Native (AIAN): A person having origins
    in any of the original peoples of North and South America (including
    Central America), and who maintains tribal affiliation or community
    attachment.

    -- Asian: A person having origins in any of the original peoples of
    the Far East, Southeast Asia, or the Indian subcontinent, including,
    for example, Cambodia, China, India, Japan, Korea, Malaysia,
    Pakistan, the Philippine Islands, Thailand and Vietnam.

    -- Black or African American: A person having origins in any of the
    black racial groups of Africa. Terms such as "Haitian" or "Negro"
    can be used in addition to "Black or African American."

    -- Native Hawaiian or Other Pacific Islander (NHOPI): A person
    having origins in any of the original peoples of Hawaii, Guam,
    Samoa, or other Pacific Islands.

    -- White: A person having origins in any of the original peoples of
    Europe, the Middle East, or North Africa

-   Ethnic Group

    -- Hispanic or Latino: A person of Cuban, Mexican, Puerto Rican,
    South or Central American or other Spanish culture or origin,
    regardless of race. The term "Spanish origin" can be used in
    addition to "Hispanic or Latino."

## Homogeneity

The hypothesis of homogeneity
$$ $$

## Individual-level data

Think of the case when a respondent is multiracial; how do you subset
the data?

Focus on results of data reduction through listwise deletion.

## Household data

Imputation and race; to show the differences when head of household is a
different race than the respondent.

## Complex survey data

Using complex survey design and the various assumptions.

### Subset option

Incorrect; without the original data and only the subset data, the
standard errors will be wrong.

### Subpopulation option

Using a subpopulation function to measure statistics and standard
errors.

## Data

We use two data sets to examine subpopulation analysis in complex survey
data. The first set of examples come from the 2009 *High School
Longitudinal Study* (HSLS:09), which initially surveyed respondents in
their first year of high school. The second set of examples come from
the *General Social Survey* (GSS), which collects data from U.S.
citizens every two years.

```{r cars}
summary(cars)
```

### Data

#### High School Longitudinal Survey (HSLS)

```{r hsls}
#| my-chunk, echo = FALSE, fig.width = 10,
#| fig.cap = "This is a long long
#|   long long caption."

plot(cars)
```

#### General Social Survey (GSS)

```{r gss}
#| echo = FALSE, fig.width = 10,
#| fig.cap = "This is a long long
#|   long long caption."

plot(cars)
```

## Subpopulation estimates

### Ratio estimators

```{r ratio-estimators, echo=FALSE}
# https://cran.r-project.org/web/packages/survey/vignettes/domain.pdf
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to
prevent printing of the R code that generated the plot.

### Sampling weights for subpopulations

We can use the `subset()` function in the *survey* `package` to handle
sub-populations. Other analysts have suggested that the `lavaan.survey`
uses the survey package to get various metrics about the population. Of
interest here are the standard errors for the population matrix.

# Results

### Base models

### Latent variable models

### Time-series models

# Discussion

# Conclusion

# References

<https://www.cdc.gov/nchs/data/natmedcare/nmc_a_04acc.pdf>

Horton N. J., and Kleinman K. P. 2007. Much ado about nothing: A
comparison of missing data methods and software to fit incomplete data
regression models. American Statistician 61: 79–90.

Lee E. S., and Forthofer R. N. 2005. Analyzing Complex Survey Data. 2nd
ed. Thousand Oaks, CA: Sage.

West, B. T., Berglund, P., & Heeringa, S. G. (2008). A Closer
Examination of Subpopulation Analysis of Complex-Sample Survey Data. The
Stata Journal, 8(4), 520-531.
<https://doi.org/10.1177/1536867X0800800404>

<https://stats.oarc.ucla.edu/r/seminars/survey-data-analysis-with-r/>

<https://doh.wa.gov/sites/default/files/legacy/Documents/1500/RaceEthnGuidelines.pdf>

<https://cran.r-project.org/web/packages/survey/vignettes/domain.pdf>

<https://www.bookdown.org/rwnahhas/RMPH/survey-domain.html>

[1]
<https://www.researchgate.net/publication/316865660_Introduction_to_the_Design_and_Analysis_of_Complex_Survey_Data>
[2]
<https://stats.oarc.ucla.edu/stata/seminars/survey-data-analysis-in-stata-17/>
[3] <https://faculty.washington.edu/tlumley/survey-jsm-nup.pdf> [4]
<https://academic.oup.com/jssam/article/11/4/743/7136601> [5]
<https://doh.wa.gov/sites/default/files/legacy/Documents/1500/RaceEthnGuidelines.pdf>
