---
title: "IMMERSE Training Notes"
author: "Nathan Alexander"
institute: 
  - "Howard University"
  - "Center for Applied Data Science and Analytics"
format: html
---


## In-person Training Day 1

### Goals for today
  1. Set up some foundations to understand latent variable modeling
  2. Explore some data in lab
  
What is a latent variable? What is a manifest or indicator variable?

  - A construct that we believe exists but we cannot directly measure; it is a *manifestation* of a set of observed variables, or indicators.

What is a latent variable (LV) measurement model?

  - Parameterizing the actual underlying relationship between the latent and observed variables.
  
How do we select indicator variables for a latent variable measurement model?

  - We look for items in our data set that we can justify as measures of that latent class variable.
  - Our latent variable is whatever it is measure by NOT only how we want to interpret the variable.

How might data equity figure into indicator selection?


Despite us carefully thinking about the content and domain of our models, we need to think about strategies to help balance information.

Where a factor analysis asks "Which items *hang* together?" we are asking different questions here. If we have a larger number of indicators which strong correlations, the latent class variables is trying to explain as much shared variance as possible.

## Pre-training Day 1

No relevant coding notes.

## Pre-training Day 2

### Set-up

```{r}
#| echo: false
#| output: false
#| warning: false

# install.packages("tidyverse", repos = "http://cran.us.r-project.org")
# install.packages("palmerpenguins", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(palmerpenguins)
library(ggplot2)
data(penguins) # load data
df <- penguins
```

#### Load palmer penguins data

```{r}
#| echo: true
#| output: true
#| warning: false
glimpse(df) # get a glimpse of the data
dim(df) # gather the dimensions of the data
names(df) # gather names of variables
summary(df) # gather a summary of the data

```

#### Create a graph using ggplot

Steps to create plot

  1. ggplot()
  
  2. Data, variable `x`, and variable `y`
  
  3. What type of graph?
  
```{r}
#| echo: true
#| output: true
#| warning: false

ggplot(data = df, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() -> plot1
plot1

```

## Pre-training Day 3

No relevant coding notes. Focus was on `dplyr` and `tidyverse` with Palmer Penguins.

## Pre-training Day 4

### Working with Mplus

Please see notes in forked files.

Please clean HSLS data.

## Pre-training Day 5

A few reminders:

  - Any variables with arrows pointing to them are dependent or endogenous variables.

  - Any variable that is independent is an exogenous variable.

  - Do not identify categorial independent/exogenous variables as `categorial` in the variable command in MPlus.

Categorical coding

Binary observed variable with LRV parameterization

