---
title: "IMMERSE Training Notes"
author: "Nathan Alexander"
institute: 
  - "Howard University"
  - "Center for Applied Data Science and Analytics"
format: html
---

## In-person Training Day 1

### Goals for today
  1. Set up some foundations to understand latent variable modeling
  2. Explore some data in lab
  
What is a latent variable? What is a manifest or indicator variable?

  - A construct that we believe exists but we cannot directly measure; it is a *manifestation* of a set of observed variables, or indicators.

What is a latent variable (LV) measurement model?

  - Parameterizing the actual underlying relationship between the latent and observed variables.
  
How do we select indicator variables for a latent variable measurement model?

  - We look for items in our data set that we can justify as measures of that latent class variable.
  - Our latent variable is whatever it is measure by NOT only how we want to interpret the variable.

How might data equity figure into indicator selection?


Despite us carefully thinking about the content and domain of our models, we need to think about strategies to help balance information.

Where a factor analysis asks "Which items *hang* together?" we are asking different questions here. If we have a larger number of indicators which strong correlations, the latent class variables is trying to explain as much shared variance as possible.


The classes that emerge are directly related to the items that you include.

## In-person Training Day 2

### Basic of the EM Algorithm

  - Initialization: Start with an initial set of estimates for the parameters.
  
  - E-Step (Extraction step): Given the current parameter estimates, calculate the expected values of the latent variables or unobserved data.
  
  - M-step (Maximization step): Maximize the expected log-likelihood function by updating the parameter estimates based on the exepcted values obtained in the E-step.
  
  - Iteration: Repeat steps 2 and 3 until convergence, which is typically determined by a predefined convergence criterion. 

*Goal is to find the parameter values that maximize the likelihood; the estimation "crawls along" with estimation to find "the peak."*

  - Stopping criteria is needed or it will iterate forever and you will never get an output file.

### Details of the modeling process

#### Class enumeration

  - Random starts note, if you do 50 random starts, you are estimating 50 mixture models. This is not necessarily what you want to do. See notes slides on Random Starts (EM Step 1).
  
  -- Look-up perturbations in relation to the modeling process.
  
  - In the output command, when you ask for `S` values, you have to ask for it, it is not automatically produced. An entire model statement is produced that cannot be directly placed into an output file; it is a model statement with start values included. *Place an asterisk with a number after it.* The @ symbol holds the start value.
  
  -- Increasing model complexity is when you see things break down. Empirical identification is the ultimate goal; we should not give it up if it does not converge the first time, we'll have to help it move along.
  
#### Deciding on the number of latent classes

How do we decide on the number of classes?
  
    - The number of classes is generally not know.
    
    - We rarely have a specific typology of how many classes.
  
## Pre-training Day 1

No relevant coding notes.

## Pre-training Day 2

### Set-up

```{r}
#| echo: false
#| output: false
#| warning: false

# install.packages("tidyverse", repos = "http://cran.us.r-project.org")
# install.packages("palmerpenguins", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(palmerpenguins)
library(ggplot2)
data(penguins) # load data
df <- penguins
```

#### Load palmer penguins data

```{r}
#| echo: true
#| output: true
#| warning: false
glimpse(df) # get a glimpse of the data
dim(df) # gather the dimensions of the data
names(df) # gather names of variables
summary(df) # gather a summary of the data

```

#### Create a graph using ggplot

Steps to create plot

  1. ggplot()
  
  2. Data, variable `x`, and variable `y`
  
  3. What type of graph?
  
```{r}
#| echo: true
#| output: true
#| warning: false

ggplot(data = df, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() -> plot1
plot1

```

## Pre-training Day 3

No relevant coding notes. Focus was on `dplyr` and `tidyverse` with Palmer Penguins.

## Pre-training Day 4

### Working with Mplus

Please see notes in forked files.

Please clean HSLS data.

## Pre-training Day 5

A few reminders:

  - Any variables with arrows pointing to them are dependent or endogenous variables.

  - Any variable that is independent is an exogenous variable.

  - Do not identify categorial independent/exogenous variables as `categorial` in the variable command in MPlus.

Categorical coding

Binary observed variable with LRV parameterization

