---
title: "REIMS sample analysis"
author: "Nathan Alexander"
date: "2024-06-03"
output:
  html_document:
    theme: flatly
  pdf_document: default
editor_options:
  markdown:
    wrap: sentence
---

# Set up

### Set up `MPlusAutomation`

```{r, eval=F}
install.packages("devtools")
library(devtools)

install_github("michaelhallquist/MplusAutomation")
```

### Load packages

```{r, echo=F, eval=T, warning=F}
library(MplusAutomation)
library(tidyverse) #collection of R packages designed for data science
library(here) #helps with filepaths
library(gt) # creates tables
library(DiagrammeR) # create path diagrams
here::i_am("reims-analysis.Rmd")
```

### Data

```{r, echo=T, eval=T, warning=F}
# set data
reims <- read_csv(here("data", "reims_clean.csv"))
```


```{r, echo=T, eval=T, warning=F}
# inspect data
reims
summary(reims)
```

#### Make updates/edits to data

We observe some of the other tags on our data. We also look at the distribution of some values. Take note that we changed the variable names (thanks Dina!) to less than eight characters so that our model could run.

```{r, echo=T, eval=T, warning=F}
# view distribution of indicator variables
table(reims$groupflag)
hist(reims$age)
table(reims$sex)
table(reims$race)

# subset data
reims1 <- reims %>% 
  select(mathperson1, mathperson2, mathperson3, mathperson4, dislikemathclass, pursuestem, boysbetter) %>% 
  rename(m1 = mathperson1,
         m2 = mathperson2,
         m3 = mathperson3,
         m4 = mathperson4,
         dislike = dislikemathclass,
         pursue = pursuestem,
         boys = boysbetter)

head(reims1, n=10)

## add ids and covariates; tell mplus that what we are

# use the reorder function to get hte variables that you want to model in the output.

```

# Models

## Model 1

```{r, echo=F, eval=T, fig.align='center'}

grViz(" digraph model {

# The `graph` statement - No editing needed

    graph [layout = dot, overlap = true]
 
# Two `node` statements
 
# One for measured variables (box) 

    node [shape=box]
    mathperson1 mathperson2 mathperson3 mathperson4;
 
# One for latent variables (circle) 
 
    node [shape=circle]
    lca [label=<Mathematics Identity <br/>C<sub>k</sub>>];
    
# `edge` statements
 
    edge [minlen = 2]
    lca -> {mathperson1 mathperson2 mathperson3 mathperson4}
 
 }") 
```

### Model 1 MplusAutomation code

```{r, eval = T}

input <- mplusObject(
  TITLE = "REIMS Mathematics Identity Model 1",
  VARIABLE = "categorical = m1 m2 m3 m4;
  usevar = m1-m4;
  classes = c(3);",
  
  ANALYSIS =
    "estimator = mlr;
    type = mixture;",
  
  OUTPUT = "tech11 tech14;",
  
  PLOT = "type = plot3;
    series = m1-m4(*);",
  
  usevariables = colnames(reims1),
  rdata = reims1)

output <- mplusModeler(input,
                       dataout = here("mplus", "reims1.dat"),
                       modelout = here("mplus", "reims1.inp"),
                       check = T, run = T, hashfilename = F)

```

Take a look at the item probability plot:

```{r}
source(here("plot_lca.txt")) # custom function created by Dina to plot our lsa output

model1 <- readModels(here("mplus", "reims1.out")) # read in output

plot_lca(model_name = model1) # there is an error with the non atomic measure columns
``` 

Show probability plot of data and observe the different classes.

---

## Model 2

Let's run a four class model and add three variables.

```{r, echo=F, eval=T, fig.align='center'}

grViz(" digraph model {

# The `graph` statement - No editing needed

    graph [layout = dot, overlap = true]
 
# Two `node` statements
 
# One for measured variables (box) 

    node [shape=box]
    mathperson1 mathperson2 mathperson3 mathperson4 dislikemathclass pursuestem boysbetter;
 
# One for latent variables (circle) 
 
    node [shape=circle]
    lca [label=<Mathematics Identity <br/>C<sub>k</sub>>];
    
# `edge` statements
 
    edge [minlen = 2]
    lca -> {mathperson1 mathperson2 mathperson3 mathperson4 dislikemathclass pursuestem boysbetter}
 
 }") 
```

### Model 2 MplusAutomation code

```{r, eval = T}

input <- mplusObject(
  TITLE = "REIMS Mathematics Identity Model 2",
  VARIABLE = "categorical = m1 m2 m3 m4 dislike pursue boys;
  usevar = m1-boys;
  classes = c(4);",
  
  ANALYSIS =
    "estimator = mlr;
    type = mixture;",
  
  OUTPUT = "tech11 tech14;",
  
  PLOT = "type = plot3;
    series = m1-boys(*);",
  
  usevariables = colnames(reims1),
  rdata = reims1)

output <- mplusModeler(input,
                       dataout = here("mplus", "reims1.dat"),
                       modelout = here("mplus", "reims1.inp"),
                       check = T, run = T, hashfilename = F)

```

Take a look at the item probability plot:

```{r}
source(here("plot_lca.txt")) # custom function created by Dina to plot our lsa output

model2 <- readModels(here("mplus", "reims1.out")) # read in output

plot_lca(model_name = model2) # there is an error with the non atomic measure columns
``` 

Show probability plot of data and observe the different classes.

---

# Enumeration

We use the mplusObject function in the MPlusAutomation package and saves all models run.

```{r}
# add new libraries
library(cowplot)
library(glue)
```

Proporion of indicators using R:

```{r}

# set up data to find proportions of binary indicators
df <- reims1 %>% 
  pivot_longer(c(m1, m2, m3, m4, dislike, pursue, boys),
  names_to = "Variable")

# create table of variables and counts
t1 <- table(df$Variable, df$value)

# find proportions and round to 3 decimal places
prop <- prop.table(t1, margin =1) %>% 
  round(3)

# combine everything to one table
dframe <- data.frame(Variables=rownames(t1), Proportion=prop[,2], Count=t1[,2])

# remove row names
row.names(dframe) <- NULL

# Make it a gt() table
prop_table <- dframe %>% 
  gt()
prop_table

# save as a word doc
gtsave(prop_table, here("figures", "prop_table.docx"))
```

Use an enumeration function

```{r, eval = F}
lca_4 <- lapply(1:4, function(k) {
  lca_enum <- mplusObject(
    
    TITLE = glue("{k}-Class"),
    
    VARIABLE = glue(
      "categorical m1-boys;
      usevar = m1-boys;
      classes = c({k});"),
    
    ANALYSIS =
      "estimator = mlr;
      type = mixture;
      starts = 500 100;",
    
    OUTPUT = "tech11 tech14 svalues;",
    
    usevariables = colnames(reims1),
    rdata = reims1)
  
  lca_enum_fit <- mplusModeler(lca_enum,
                               dataout = glue(here("enum", "reims1.dat")),
                               modelout = glue(here("enum", "c{k}_reims1.inp")),
                               check = T, run = T, hashfilename = F)})
```

### table of fit

We want to begin by extracting the data:

```{r}
output_reims1 <- readModels(here("enum"))

enum_extract <- LatexSummaryTable(
  output_reims1,
  keepCols = c(
    "Title",
    "Parameters",
    "LL",
    "BIC",
    "aBIC",
    "BLRT_PValue",
    "T11_VLMR_PValue",
    "Observations"
  ),
  sortBy = "Title"
) # select first set of models (Class 1 through 4)

allFit <- enum_extract %>% 
  mutate(CAIC = -2 * LL + Parameters * (log(Observations) + 1)) %>% 
  mutate(AWE = -2 * LL + 2 * Parameters * (log(Observations) + 1.5)) %>% 
  mutate(SIC = -.5 * BIC) %>% 
  mutate(expSIC = exp(SIC - max(SIC))) %>% 
  mutate(BF = exp(SIC - lead(SIC))) %>% 
  mutate(cmPk = expSIC / sum(expSIC)) %>% 
  dplyr::select(1:5, 9:10, 6:7, 13, 14) %>%
  arrange(Parameters)
```

Then we create a table:

```{r}
fit_table <- allFit %>%
  gt() %>%
  tab_header(title = md("**Model Fit Summary Table**")) %>%
  cols_label(
    Title = "Classes",
    Parameters = md("Par"),
    LL = md("*LL*"),
    T11_VLMR_PValue = "VLMR",
    BLRT_PValue = "BLRT",
    BF = md("BF"),
    cmPk = md("*cmPk*")
  ) %>%
  tab_footnote(
    footnote = md(
      "*Note.* Par = Parameters; *LL* = model log likelihood;
BIC = Bayesian information criterion;
aBIC = sample size adjusted BIC; CAIC = consistent Akaike information criterion;
AWE = approximate weight of evidence criterion;
BLRT = bootstrapped likelihood ratio test p-value;
VLMR = Vuong-Lo-Mendell-Rubin adjusted likelihood ratio test p-value;
*cmPk* = approximate correct model probability."
    ),
locations = cells_title()
  ) %>%
  tab_options(column_labels.font.weight = "bold") %>%
  fmt_number(c(3:7),
             decimals = 2) %>%
  sub_missing(1:11,
              missing_text = "--") %>%
  fmt(
    c(8:9, 11),
    fns = function(x)
      ifelse(x < 0.001, "<.001",
             scales::number(x, accuracy = .01))
  ) %>%
  fmt(
    10,
    fns = function (x)
      ifelse(x > 100, ">100",
             scales::number(x, accuracy = .01))
  ) %>%  
  tab_style(
    style = list(
      cell_text(weight = "bold")

            ),
    locations = list(cells_body(
     columns = BIC,
     row = BIC == min(BIC[1:6]) # Change this to the number of classes you estimated
    ),
    cells_body(
     columns = aBIC,
     row = aBIC == min(aBIC[1:6])
    ),
    cells_body(
     columns = CAIC,
     row = CAIC == min(CAIC[1:6])
    ),
    cells_body(
     columns = AWE,
     row = AWE == min(AWE[1:6])
    ),
    cells_body(
     columns = cmPk,
     row =  cmPk == max(cmPk[1:6])
     ),    
    cells_body(
     columns = BF,
     row =  BF > 10),
    cells_body( 
     columns =  T11_VLMR_PValue,
     row =  ifelse(T11_VLMR_PValue < .001 & lead(T11_VLMR_PValue) > .05, T11_VLMR_PValue < .001, NA)),
    cells_body(
     columns =  BLRT_PValue,
     row =  ifelse(BLRT_PValue < .001 & lead(BLRT_PValue) > .05, BLRT_PValue < .001, NA))
  )
) 

fit_table

```

save the table:

```{r}
gtsave(fit_table, here("figures", "fit_table.png"))
```

### Information Criterion Plot

```{r height=5, width=7}
allFit %>%
  dplyr::select(2:7) %>%
  rowid_to_column() %>%
  pivot_longer(`BIC`:`AWE`,
               names_to = "Index",
               values_to = "ic_value") %>%
  mutate(Index = factor(Index,
                        levels = c ("AWE", "CAIC", "BIC", "aBIC"))) %>%
  ggplot(aes(
    x = rowid,
    y = ic_value,
    color = Index,
    shape = Index,
    group = Index,
    lty = Index
  )) +
  geom_point(size = 2.0) + geom_line(size = .8) +
  scale_x_continuous(breaks = 1:nrow(allFit)) +
  scale_colour_grey(end = .5) +
  theme_cowplot() +
  labs(x = "Number of Classes", y = "Information Criteria Value", title = "Information Criteria") +
  theme(
    text = element_text(family = "serif", size = 12),
    legend.text = element_text(family="serif", size=12),
    legend.key.width = unit(3, "line"),
    legend.title = element_blank(),
    legend.position = "top"  
  )
```

save the figure:

```{r}
ggsave(here("figures", "info_criteria.png"), dpi=300, height=5, width=7, units="in")
```

### Compare class solutions

Compare probability plots for $K = 1:4$ class solutions

```{r}
model_results <- data.frame()

for (i in 1:length(output_reims1)) {
  temp <- output_reims1[[i]]$parameters$probability.scale %>% 
    mutate(model = paste0(i, "-Class Model"))
  
  model_results <- rbind(model_results, temp)
}

compare_plot <-
  model_results %>% 
  filter(category ==2) %>% 
  dplyr::select(est, model, LatentClass, param)

compare_plot$param <- fct_inorder(compare_plot$param)

ggplot(
  compare_plot,
  aes(
    x=param,
    y=est,
    color = LatentClass,
    shape = LatentClass,
    group = LatentClass,
    lty = LatentClass
  )
) +
  geom_point() +
  geom_line() +
  scale_color_viridis_d() +
  facet_wrap(~ model, ncol = 2) +
  labs(title = "Mathematics Identity Items", x = " ", y = "Probability") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        axis.text.x = element_text(angle = -45, hjust = -.1))

```

save the figure:

```{r, eval = FALSE}
ggsave(here("figures", "compare_kclass_plot.png"), dpi=300, height=5, width=7, units="in")
```


### 4-Class Probability Plot

Use the `plot_lca` function provided in the folder to plot the item probability plot. This function requires one argument:
- `model_name`: The name of the Mplus `readModels` object (e.g., `output_lsal$c4_lsal.out`) - this was updated for reims.

```{r fig.height=6, fig.width=10}
source("plot_lca.txt")

plot_lca(model_name = output_reims1$c4_reims1.out)
```

save the figure:

```{r, eval = FALSE}
ggsave(here("figures", "probability_plot_4class.png"), dpi="retina", height=5, width=7, units="in")
```

### 2-Class Probability Plot

```{r fig.height=6, fig.width=10}

source("plot_lca.txt")

plot_lca(model_name = output_reims1$c2_reims1.out)

```

save the figure:

```{r, eval = FALSE}
ggsave(here("figures", "probability_plot_2class.png"), dpi="retina", height=5, width=7, units="in")
```

### Observed response patterns

Save response frequencies for the 2-class model with `response is _____.dat` under `SAVEDATA`.

```{r, eval = FALSE}

patterns  <- mplusObject(
  
  TITLE = "LCA - Save response patterns", 
  
  VARIABLE = 
  "categorical = m1-boys; 
   usevar =  m1-boys;
   classes = c(4);",
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture;    
    starts = 0; 
    processors = 10;
    optseed = 830529;",
  
  SAVEDATA = 
   "File=savedata.dat;
    Save=cprob;
    
    ! Code to save response frequency data 
    
    response is resp_patterns.dat;",
  
  OUTPUT = "patterns tech10 tech11 tech14",
  
  usevariables = colnames(reims1),
  rdata = reims1)

patterns_fit <- mplusModeler(patterns,
                dataout=here("mplus", "patterns.dat"),
                modelout=here("mplus", "patterns.inp") ,
                check=TRUE, run = TRUE, hashfilename = FALSE)
```

read in observed response pattern data and relabel the columns:

```{r}

# read in response frequency data that we just created:
patterns <- read_table(here("mplus", "resp_patterns.dat"),
                        col_names=FALSE, na = "*") 

# extract the column names
names <- names(readModels(here("mplus", "patterns.out"))[['savedata']])

# add the names back to the dataset
colnames(patterns) <- c("Frequency", names)  

```

we then create a table with the response patterns, then top of conditional response pattern for each modal class assignment

```{r,eval=TRUE}
# Order responses by highest frequency
order_highest <- patterns %>% 
  arrange(desc(Frequency)) 

# Loop `patterns` data to list top 5 conditional response patterns for each class
loop_cond  <- lapply(1:max(patterns$C), function(k) {       
order_cond <- patterns %>%                    
  filter(C == k) %>%                    
  arrange(desc(Frequency)) %>%                
  head(5)                                     
  })                                          

# Convert loop into data frame
table_data <- as.data.frame(bind_rows(loop_cond))

# Combine unconditional and conditional responses patterns
response_patterns <-  rbind(order_highest[1:5,], table_data) 
```

we then use `{gt}` to make a nicely formatted table.

```{r}
resp_table <- response_patterns %>% 
  gt() %>%
    tab_header(
    title = "Observed Response Patterns",
    subtitle = html("Response patterns, estimated frequencies, estimated posterior class probabilities and modal assignments")) %>% 
    tab_source_note(
    source_note = md("Data Source: **Racial Ethnic Identity in Mathematics Survey (REIMS)**")) %>%
    cols_label(
      Frequency = html("<i>f</i><sub>r</sub>"),
    M1 = "Math Person 1",
    M2 = "Math Person 2",
    M3 = "Math Person 3",
    M4 = "Math Person 4",
    DISLIKE = "Dislike Math",
    PURSUE = "Pursue Math",
    BOYS = "Boys Better",
    CPROB1 = html("P<sub><i>k</i></sub>=1"),
    CPROB2 = html("P<sub><i>k</i></sub>=2"),
    CPROB3 = html("P<sub><i>k</i></sub>=3"),
    CPROB4 = html("P<sub><i>k</i></sub>=4"),    # Change based on number of classes
    C = md("*k*")) %>% 
  tab_row_group(
    label = "Unconditional response patterns",
    rows = 1:5) %>%
  tab_row_group(
    label = md("*k* = 1 Conditional response patterns"),
    rows = 6:10) %>% #EDIT THESE VALUES BASED ON THE LAST COLUMN
  tab_row_group(
    label = md("*k* = 2 Conditional response patterns"),
    rows = 11:15)  %>% #EDIT THESE VALUES BASED ON THE LAST COLUMN
  tab_row_group(
    label = md("*k* = 3 Conditional response patterns"),
    rows = 16)  %>% #EDIT THESE VALUES BASED ON THE LAST COLUMN
  tab_row_group(
    label = md("*k* = 4 Conditional response patterns"),
    rows = 17:21)  %>% #EDIT THESE VALUES BASED ON THE LAST COLUMN  
    row_group_order(
      groups = c("Unconditional response patterns",
                 md("*k* = 1 Conditional response patterns"),
                 md("*k* = 2 Conditional response patterns"),
                 md("*k* = 3 Conditional response patterns"),
                 md("*k* = 4 Conditional response patterns"))) %>% 
    tab_footnote(
    footnote = html(
      "<i>Note.</i> <i>f</i><sub>r</sub> = response pattern frequency; P<sub><i>k</i></sub> = posterior class probabilities"
    )
  ) %>% 
  cols_align(align = "center") %>% 
  opt_align_table_header(align = "left") %>% 
  gt::tab_options(table.font.names = "Times New Roman") 

resp_table
```

save the table

```{r}
gtsave(resp_table, here("figures", "resp_table.png"))
```


### Classification Diagnostics

We will use Mplus to calculate k-class confidence intervals, using the 4-class model.

```{r, eval = F}

classification <- mplusObject(
  
  TITLE = "LCA - Calculate k-Class 95% CI",
  
  VARIABLE =
    "categorical = m1-boys;
    usevar = m1-boys;
    classes = c(4);",
  
  ANALYSIS =
    "estimator = ml;
    type = mixture;
    starts = 0;
    processors = 10;
    optseed = 945065;
    bootstrap = 100;",
  
  MODEL =
    "
  !CHANGE THIS SECTION TO YOUR CHOSEN k-CLASS MODEL
  
  %OVERALL%
  [C#1](c1);
  [C#2](c2);
  [C#3](c3);


  Model Constraint:
  New(p1 p2 p3 p4);
  
  p1 = exp(c1)/(1+exp(c1)+exp(c2)+exp(c3));
  p2 = exp(c2)/(1+exp(c1)+exp(c2)+exp(c3));
  p3 = exp(c3)/(1+exp(c1)+exp(c2)+exp(c3));  
  p4 = 1/(1+exp(c1)+exp(c2)+exp(c3));",
  
  OUTPUT = "cinterval(bcbootstrap)",
  
  usevariables = colnames(reims1),
  rdata = reims1)

classification_fit <- mplusModeler(classification,
                dataout=here("mplus", "reims-1.dat"),
                modelout=here("mplus", "class.inp") ,
                check=TRUE, run = TRUE, hashfilename = FALSE)
```

*Note from IMMERSE team*: Ensure that the classes did not shift during this step (i.g., Class 1 in the enumeration run is now Class 4). Evaluate output and compare the class counts and proportions for the latent classes. Using the OPTSEED function ensures replication of the best loglikelihood value run.  

---

Read in the 4-class model:

```{r}
# Read in the 4-class model and extract information needed
class_output <- readModels(here("mplus", "class.out"))

# Entropy
entropy <- c(class_output$summaries$Entropy, rep(NA, class_output$summaries$NLatentClasses-1))

# 95% k-Class and k-class 95% Confidence Intervals
k_ci <- class_output$parameters$ci.unstandardized %>% 
  filter(paramHeader == "New.Additional.Parameters") %>% 
  unite(CI, c(low2.5,up2.5), sep=", ", remove = TRUE) %>% 
  mutate(CI = paste0("[", CI, "]")) %>% 
  rename(kclass=est) %>% 
  dplyr::select(kclass, CI)

# AvePPk = Average Latent Class Probabilities for Most Likely Latent Class Membership (Row) by Latent Class (Column)
avePPk <- tibble(avePPk = diag(class_output$class_counts$avgProbs.mostLikely))

# mcaPk = modal class assignment proportion 
mcaPk <- round(class_output$class_counts$mostLikely,3) %>% 
  mutate(model = paste0("Class ", class)) %>% 
  add_column(avePPk, k_ci) %>% 
  rename(mcaPk = proportion) %>% 
  dplyr::select(model, kclass, CI, mcaPk, avePPk)

# OCCk = odds of correct classification
OCCk <- mcaPk %>% 
  mutate(OCCk = round((avePPk/(1-avePPk))/(kclass/(1-kclass)),3))

# Put everything together
class_df <- data.frame(OCCk, entropy)
```

now we use `{gt}` to make a nicely formatted table

```{r}
class_table <- class_df %>% 
  gt() %>%
    tab_header(
    title = "Model Classification Diagnostics for the 4-Class Solution") %>%
    cols_label(
      model = md("*k*-Class"),
      kclass = md("*k*-Class Proportions"),
      CI = "95% CI",
      mcaPk = md("*mcaPk*"),
      avePPk = md("*AvePPk*"),
      OCCk = md("*OCCk*"),
      entropy = "Entropy") %>% 
    sub_missing(7,
              missing_text = "") %>%
  cols_align(align = "center") %>% 
  opt_align_table_header(align = "left") %>% 
  gt::tab_options(table.font.names = "Times New Roman")

class_table
```

save the table:

```{r}
gtsave(class_table, here("figures", "class_table.png"))
```

---

# Auxillary variables

## Model 3

We will now add auxiliary variables to our model.

```{r, echo=F, eval=T, fig.align='center'}

grViz(" digraph model {

# The `graph` statement - No editing needed

    graph [layout = dot, overlap = true]
 
# Two `node` statements
 
# One for measured variables (box) 

    node [shape=box]
    mathperson1 mathperson2 mathperson3 mathperson4 dislikemathclass pursuestem boysbetter;
 
# One for latent variables (circle) 
 
    node [shape=circle]
    lca [label=<Mathematics Identity <br/>C<sub>k</sub>>];
    
# `edge` statements
 
    edge [minlen = 2]
    lca -> {mathperson1 mathperson2 mathperson3 mathperson4 dislikemathclass pursuestem boysbetter}
    
    node [shape=box]
    gender -> lca ;
    age -> lca;
    race -> lca;
 
 }") 
```

---
This file is based on resources provided by the Institute of Mixture Modeling for Equity-Oriented Researchers, Scholars, and Educators (2024). IMMERSE In-Person Training Workshop (IES No. 305B220021).
Institute of Education Sciences.
<https://immerse-ucsb.github.io/pre-training>